---
title: "Modeling expression changes at gene level"
author: "Migdal"
date: "2/15/2022"
output:
  html_document:
    self_contained: yes
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

# Introduction

This vignette extends the 
[xcore user guide](https://bkaczkowski.github.io/xcore/articles/xcore_vignette.html) 
by showing how to perform gene level expression modeling using RNA-seq and 
microarray input data.

In *xcore* we model the expression as a function of promoter's sequence 
described by the ChIP-seq signatures. Similarly to differential expression
analysis, one can consider the expression at the level of genes or transcripts.
While most of the analysis steps would be identical for the two variants, the 
key difference is in the choice of promoters annotation. Particularly, different 
gene's transcripts can have distinct promoters. One the other hand, expression 
at gene level is an aggregation of transcripts expression yet we would rather 
like to use only one promoter to describe a single gene.

*xcoredata* provides ChIP-seq based molecular signatures constructed against
FANTOM5 promoters. These promoters correspond to transcripts' promoters making
them very easy to use with CAGE data mapped to FANTOM5 reference (as exemplified 
in 
[xcore user guide](https://bkaczkowski.github.io/xcore/articles/xcore_vignette.html)). 
Additionally, *xcoredata* provides a set of core FANTOM5 promoters 
(`promoters_f5_core`) which defines only one promoter per gene. Using this core 
set of promoters it is possible to construct a gene level analysis with 
non-CAGE data, by matching input gene identifiers with core FANTOM5 promoters.

The reminder of this document shows how to construct a gene level *xcore* 
analysis. For this purpose we use publicly available RNA-seq time-series dataset 
obtained from classic TGFꞵ induced epithelial-to-mesenchymal transition 
experiment conducted in A-549 cell line 
([Chang H, et al. NAR. 2016.](https://doi.org/10.1093/nar/gkw126), [GSE69667](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE69667)).
The other example uses microarray time-series dataset also from TGFꞵ induced 
epithelial-to-mesenchymal transition experiment conducted in A-549 cell line 
([Sartor MA, et al. Bioinformatics. 2010.](https://pubmed.ncbi.nlm.nih.gov/20007254/), 
[GSE17708](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE17708)).

# RNA-seq data
## Obtaining the expression data

We have downloaded the raw counts and metadata for GSE69667 study from 
[GREIN](http://www.ilincs.org/apps/grein/?gse=GSE69667) portal that provide raw 
counts for many RNA-seq datasets deposited in the GEO repository. Before these
can be used in *xcore* framework, downloaded metadata needs to be combined with 
raw counts.

If you are interested in the processing details, look up the Rmd source file
of this vignette.

```{r eval=FALSE, echo=FALSE}
counts_rna_seq <- read.csv("GSE69667_GeneLevel_Raw_data.csv", row.names = "X")
gene_symbol <- counts_rna_seq$gene_symbol
i_dups <- duplicated(gene_symbol) | duplicated(gene_symbol, fromLast = TRUE)
counts_rna_seq <- counts_rna_seq[! i_dups, colnames(counts_rna_seq) != "gene_symbol"] %>%
  as.matrix() %>%
  round()
rownames(counts_rna_seq) <- gene_symbol[! i_dups]
mode(counts_rna_seq) <- "integer"
metadata <- read.csv("GSE69667_metadata.csv", row.names = "X")
colnames(counts_rna_seq) <-
  sub(pattern = " ",
      replacement = "_",
      x = metadata[colnames(counts_rna_seq), "characteristics"])

out_prefix <- system.file("inst", "extdata", package = "xcore")
save(counts_rna_seq, file = file.path(out_prefix, "GSE69667.rda"))
```

Here we recommend that ready to use counts matrix can be downloaded from our 
server.

```{r message=FALSE}
download.file(url = "https://zdglab.iimcb.gov.pl/mmigdal/GSE69667.rda", 
              destfile = "GSE69667.rda")
load("GSE69667.rda")
```

## Constructing experiment design

The code below shows how to construct design matrix, by taking advantage of the
patterns present in the sample names.

```{r}
knitr::kable(head(counts_rna_seq))
```

```{r}
cond <- sub(pattern = "_.*", replacement = "", x = colnames(counts_rna_seq))
ncond <- cond %>% unique() %>% length()
design <- diag(ncond)[rep(seq_len(ncond), times = table(cond)[unique(cond)]), ]
rownames(design) <- colnames(counts_rna_seq)
colnames(design) <- unique(cond)

knitr::kable(design)
```

## Obtaining core promoters annotation and molecular signatures 

```{r message=FALSE}
promoters_f5_core <- xcoredata::promoters_f5_core()
remap_promoters_f5 <- xcoredata::remap_promoters_f5()
```

## Translating gene symbols into FANTOM5 core promoters

This is the key step in gene level analysis. At first we construct a named 
character vector mapping gene symbols to FANTOM5 core promoters. Then we can use
it to translate our counts matrix using `translateCounts` function.

```{r}
symbol2fantom <- setNames(promoters_f5_core$name, promoters_f5_core$SYMBOL)
counts_rna_seq_fantom <- translateCounts(counts_rna_seq, dict = symbol2fantom)
```

## Processing the counts and molecular signatures

```{r}
mae_rna_seq <- prepareCountsForRegression(
  counts = counts_rna_seq_fantom,
  design = design,
  base_lvl = "0h"
)
mae_rna_seq <- addSignatures(mae_rna_seq, remap = remap_promoters_f5)
mae_rna_seq <- filterSignatures(mae_rna_seq, min = 0.05, max = 0.95)
```

## Modeling gene expression

As always, depending on infrastructure you are using this step can be time 
consuming.

```{r message=FALSE, cache=TRUE, cache.lazy=FALSE}
# register parallel backend
doMC::registerDoMC(cores = 5L)

# set seed
set.seed(314159265)

res_rna_seq <- modelGeneExpression(
  mae = mae_rna_seq,
  xnames = "remap",
  nfolds = 5)
```

## Exploring results

```{r warning=FALSE}
top_signatures <- head(res_rna_seq$results$remap, 15)
pheatmap::pheatmap(
  mat = top_signatures[, 2:8],
  labels_row = top_signatures$name,
  cluster_cols = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(35),
  breaks = seq(from = -0.1, to = 0.1, length.out = 36),
  main = "TGFꞵ EMT in A-549 cell line (GSE69667)\nReMap2020molecular signatures activity"
)
```

# Microarray data
## Obtaining the expression data

Here, we have downloaded supplementary data file containing pre-processed data 
from [GSE17708]() study. This supplementary file is an xls file that needs to
be loaded to R and converted into matrix before we can use it.

```{r eval=FALSE, echo=FALSE}
download.file(
  url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE17708&format=file&file=GSE17708%5FKeshamouni%5FTGFB1%5Flogs%2Exls%2Egz", 
  destfile = "GSE17708_Keshamouni_TGFB1_logs.xls.gz")
R.utils::gunzip("GSE17708_Keshamouni_TGFB1_logs.xls.gz", remove = TRUE)
counts_microarray <- readxl::read_excel(
  path = "GSE17708_Keshamouni_TGFB1_logs.xls", 
  col_names = TRUE, 
  skip = 4)
samples_cols <- grep(pattern = "^V[0-9]", x = colnames(counts_microarray), value = TRUE)
samples <- c(
  "0hr_1", "0hr_2", "0hr_3",       # 0hr
  "0.5hr_1", "0.5hr_2", "0.5hr_3", # 0.5hr
  "1hr_1", "1hr_2", "1hr_3",       # 1hr
  "2hr_1", "2hr_2",                # 2hr
  "4hr_1", "4hr_2", "4hr_3",       # 4hr
  "8hr_1", "8hr_2", "8hr_3",       # 8hr
  "16hr_1", "16hr_2", "16hr_3",    # 16hr
  "24hr_1", "24hr_2", "24hr_3",    # 24hr
  "72hr_1", "72hr_2", "72hr_3"     # 72hr
)
entrez_gene_id <- counts_microarray[["Entrez Gene"]]
counts_microarray <- counts_microarray %>%
  dplyr::select(setNames(object = samples_cols, nm = samples)) %>%
  as.matrix() %>%
  type.convert(as.is = TRUE) %>%
  rowsum(group = entrez_gene_id, reorder = FALSE) %>%
  round()
mode(counts_microarray) <- "integer"

out_prefix <- system.file("inst", "extdata", package = "xcore")
save(counts_microarray, file = file.path(out_prefix, "GSE17708.rda"))
```

Ready to use "counts" matrix (these are not raw counts) can be downloaded from 
our server.

```{r message=FALSE}
download.file(url = "https://zdglab.iimcb.gov.pl/mmigdal/GSE17708.rda", 
              destfile = "GSE17708.rda")
load("GSE17708.rda")
```


```{r}
knitr::kable(head(counts_microarray))
```

## Constructing experiment design

Again we construct design matrix, by taking advantage of the patterns present in 
the sample names.

```{r}
cond <- sub(pattern = "_.*", replacement = "", x = colnames(counts_microarray))
ncond <- cond %>% unique() %>% length()
design <- diag(ncond)[rep(seq_len(ncond), times = table(cond)[unique(cond)]), ]
rownames(design) <- colnames(counts_microarray)
colnames(design) <- unique(cond)

knitr::kable(design)
```

## Translating gene symbols into FANTOM5 core promoters

Next, we need to translate expression matrix's row IDs to FANTOM5 core promoters.
The only difference is that this time we map ENTREZ IDs to FANTOM5 promoters.

```{r}
# TODO move to xcoredata
entrez2fantom <- setNames(promoters_f5_core$name, promoters_f5_core$ENTREZID)
i_dups <- 
  duplicated(promoters_f5_core$ENTREZID) | duplicated(promoters_f5_core$ENTREZID, fromLast = TRUE)
entrez2fantom <- entrez2fantom[! i_dups]
counts_fantom <- translateCounts(counts_microarray, dict = entrez2fantom)
```

## Processing the counts and molecular signatures

Here, our workflow deviates from the previous example. Because, we use the 
pre-processed expression data we don't really want to normalize it. Instead
we will use the data as is.

```{r}
mae_microarray <- regressionData(
  expr_mat = counts_fantom,
  design = design,
  base_lvl = "0hr"
)
mae_microarray <- addSignatures(mae_microarray, remap = remap_promoters_f5)
mae_microarray <- filterSignatures(mae_microarray, min = 0.05, max = 0.95)
```

## Modeling gene expression

```{r message=FALSE, cache=TRUE, cache.lazy=FALSE}
# register parallel backend
doMC::registerDoMC(cores = 5L)

# set seed
set.seed(314159265)

res_microarray <- modelGeneExpression(
  mae = mae_microarray,
  xnames = "remap",
  nfolds = 5)
```

## Exploring results

```{r warning=FALSE}
top_signatures <- head(res_microarray$results$remap, 15)
pheatmap::pheatmap(
  mat = top_signatures[, 2:9],
  labels_row = top_signatures$name,
  cluster_cols = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(35),
  breaks = seq(from = -0.05, to = 0.05, length.out = 36),
  main = "TGFꞵ EMT in A-549 cell line (GSE17708)\nReMap2020 molecular signatures activity"
)
```
