---
title: "Maciej Migdał"
author: "Anonymous"
output: pdf_document
---

```{r include=FALSE}
library(magrittr)
devtools::load_all()

# Function return logical vector indicating columns populated in all rows
# mat - object of class dgCMatrix with only 1 and 0 values
getSharedColumns <- function(mat) {
    stopifnot(is(mat, "dgCMatrix"))
    stopifnot(nrow(mat) > 0)

    Matrix::colSums(mat) == nrow(mat)
}

# Funtion return features from subject overlapped by query
# where subject, query are GRanges objects
# ... other arguments internally passed to GenomicRanges::findOverlaps
intersectGr <- function(subject, query, ...) {
    stopifnot(is(subject, "GRanges"))
    stopifnot(is(query, "GRanges"))

    hits <- GenomicRanges::findOverlaps(query = query, subject = subject, ...)
    i <- S4Vectors::subjectHits(hits)
    subject[i, ]

# ENCODE blacklist list 636 regions spanning ~7% of human genome
blacklist <- rtracklayer::import(system.file("inst",
                                             "extdata",
                                             "hg38-blacklist.v2.bed.gz",
                                             package = "xcore"))
blacklisted_promoters <- intersectGr(promoters_f5, blacklist, type = "any") %>%
    GenomicRanges::mcols() %>%
    `[[`("name")

# ENSEMBL soft mask list 4725854 regions spanning ~50% of human genome
ensembl_sm <- rtracklayer::import(system.file("inst",
                                              "extdata",
                                              "hg38_ensembl_sm.bed.gz",
                                              package = "xcore"))
ensembl_sm_promoters <- intersectGr(promoters_f5, ensembl_sm, type = "any") %>%
    GenomicRanges::mcols() %>%
    `[[`("name")
}
```


# Deeper look into the “signatures”

## Technical biases and noise vs biological signal.

1. Observation made by Bogumil is that some promoters always give a signal
   regardless on the transcription factor we look at. The reverse can be also
   observed - TF gives signal regardless of the promoter we look at. The aim
   here is to show this observation.

### Looking at promoter TF interaction matrix row wise (promoter wise)

Number of promoters: `r nrow(remap_promoters)`

ReMap2020 TFs: `r ncol(remap_promoters)`

ChIP-Atlas TFs: `r ncol(chip_atlas_promoters)`

```{r promoters_coverage_remap}
remap_promoters_counts <- remap_promoters %>%
    Matrix::rowSums()

remap_promoters_counts %>%
    summary()

remap_promoters_counts %>%
    graphics::hist(breaks = 100,
                   plot = TRUE,
                   main = "ReMap2020 promoter coverage")

(remap_promoters_counts <= 50) %>% table() / length(remap_promoters_counts)

(remap_promoters_counts == 0) %>% table() / length(remap_promoters_counts)
(m <- remap_promoters_counts == 0) %>% sum()

# What genes are those 0 promoters assigned to?
zero_promoters <- names(m)[m]
zero_promoters <- promoters_f5[promoters_f5$name %in% zero_promoters, ]
zero_promoters_symbol <- zero_promoters$SYMBOL %>% table()
zero_promoters_symbol[zero_promoters_symbol > 5]

# Lets look into promoters with the highest number of transcription factors,
# how many of those are shared?
remap_promoters_highest <- remap_promoters[remap_promoters_counts >= 3250, ]
shared_cols_m <- getSharedColumns(remap_promoters_highest)
sum(shared_cols_m)
highest_promoters <- remap_promoters_highest %>% rownames()
highest_promoters
promoters_f5[promoters_f5$name %in% highest_promoters, ]$SYMBOL

# How does the number of shared TFs depends on how many top promoters we are
# looking at?
shared_tf <- c()
shared_tf_nprom <- c()
x <- seq(from = 1, to = 3322, by = 50)
for (n in x) {
    sel <- remap_promoters[remap_promoters_counts >= n, , drop = FALSE]
    shared_tf_nprom <- c(shared_tf_nprom, nrow(sel))
    n_tf <- sel %>%
        getSharedColumns() %>%
        sum()
    shared_tf <- c(shared_tf, n_tf)
}
plot(x = x,
     y = shared_tf,
     main = "ReMap2020 promoters coverage vs number of shared TFs",
     xlab = "promoter coverage >= x",
     ylab = "number of shared TFs")
plot(x = shared_tf_nprom,
     y = shared_tf,
     main = "ReMap2020 number of promoters vs number of shared TFs",
     xlab = "number of promoters",
     ylab = "number of shared TFs")

# Blacklisted promoters vs TFs coverage
blk <- remap_promoters_counts[
        names(remap_promoters_counts) %in% blacklisted_promoters]
length(blk)
nonblk <- remap_promoters_counts[
        ! names(remap_promoters_counts) %in% blacklisted_promoters]
graphics::boxplot(blk,
                  nonblk,
                  main = "Blacklisted regions vs TF count",
                  names = c("blacklisted", "non blacklisted"))

# Ensembl soft masked promoters vs TFs coverage
sm <- remap_promoters_counts[
        names(remap_promoters_counts) %in% ensembl_sm_promoters]
length(sm)
nonsm <- remap_promoters_counts[
        ! names(remap_promoters_counts) %in% ensembl_sm_promoters]
graphics::boxplot(sm,
                  nonsm,
                  main = "Ensembl soft masked regions vs TF count",
                  names = c("soft masked", "non soft masked"))
```

```{r promoters_coverage_chip_atlas}
# TODO set chip_atlas_promoters class to  dgCMatrix
chip_atlas_promoters <- as(chip_atlas_promoters, "dgCMatrix")

chip_atlas_promoters_counts <- chip_atlas_promoters %>%
    Matrix::rowSums()

chip_atlas_promoters_counts %>%
    summary()

chip_atlas_promoters_counts %>%
    graphics::hist(breaks = 100,
                   plot = TRUE,
                   main = "ChIP-Atlas promoter coverage")

(chip_atlas_promoters_counts <= 50) %>%
    table() / length(chip_atlas_promoters_counts)

(chip_atlas_promoters_counts == 0) %>%
    table() / length(chip_atlas_promoters_counts)

(m <- chip_atlas_promoters_counts == 0) %>% sum()

# What genes are those 0 promoters assigned to?
zero_promoters <- names(m)[m]
zero_promoters <- promoters_f5[promoters_f5$name %in% zero_promoters, ]
zero_promoters_symbol <- zero_promoters$SYMBOL %>% table()
zero_promoters_symbol[zero_promoters_symbol > 5]

# Lets look into promoters with the highest number of transcription factors,
# how many of those are shared?
chip_atlas_promoters_highest <-
    chip_atlas_promoters[chip_atlas_promoters_counts == 14869, ]
shared_cols_m <- getSharedColumns(chip_atlas_promoters_highest)
sum(shared_cols_m)
highest_promoters <- chip_atlas_promoters_highest %>% rownames()
highest_promoters
promoters[promoters_f5$name %in% highest_promoters, ]$SYMBOL

# How does the number of shared TFs depends on how many top promoters we are
# looking at?
shared_tf <- c()
shared_tf_nprom <- c()
x <- seq(from = 1, to = 14869, by = 50)
for (n in x) {
    sel <-
        chip_atlas_promoters[chip_atlas_promoters_counts >= n, , drop = FALSE]
    shared_tf_nprom <- c(shared_tf_nprom, nrow(sel))
    n_tf <- sel %>%
        getSharedColumns() %>%
        sum()
    shared_tf <- c(shared_tf, n_tf)
}
plot(x = x,
     y = shared_tf,
     main = "ChIP-Atlas promoters coverage vs number of shared TFs",
     xlab = "promoter coverage >= x",
     ylab = "number of shared TFs")
plot(x = shared_tf_nprom,
     y = shared_tf,
     main = "ChIP-Atlas number of promoters vs number of shared TFs",
     xlab = "number of promoters",
     ylab = "number of shared TFs")

# Blacklisted promoters vs TFs coverage
blk <- chip_atlas_promoters_counts[
        names(chip_atlas_promoters_counts) %in% blacklisted_promoters]
length(blk)
nonblk <- chip_atlas_promoters_counts[
        ! names(chip_atlas_promoters_counts) %in% blacklisted_promoters]
graphics::boxplot(blk,
                  nonblk,
                  main = "Blacklisted regions vs TF count",
                  names = c("blacklisted", "non blacklisted"))

# Ensembl soft masked promoters vs TFs coverage
sm <- chip_atlas_promoters_counts[
        names(chip_atlas_promoters_counts) %in% ensembl_sm_promoters]
length(sm)
nonsm <- chip_atlas_promoters_counts[
        ! names(chip_atlas_promoters_counts) %in% ensembl_sm_promoters]
graphics::boxplot(sm,
                  nonsm,
                  main = "Ensembl soft masked regions vs TF count",
                  names = c("soft masked", "non soft masked"))
```

### Looking at promoter TF interaction matrix column wise (TF wise)

```{r tf_coverage_remap}
remap_tf_counts <- remap_promoters %>%
    Matrix::colSums()

remap_tf_counts %>%
    summary()

remap_tf_counts %>%
    `/`(nrow(remap_promoters)) %>%
    summary()

remap_tf_counts %>%
    graphics::hist(breaks = 100,
                   plot = TRUE,
                   main = "ReMap2020 TFs coverage")

(remap_tf_counts <= 50) %>% table() / length(remap_tf_counts)

(remap_tf_counts == 0) %>% table() / length(remap_tf_counts)
(m <- remap_tf_counts == 0) %>% sum()

# How many TFs overlapp X% or more promoters
vapply(X = seq(from = 0, to = 1, by = 0.1),
       FUN = function(x) sum(remap_tf_counts >= (nrow(remap_promoters) * x)),
       FUN.VALUE = numeric(1L)) %>%
    stats::setNames(nm = seq(from = 0, to = 100, by = 10))
# what are the top TFs?
remap_tf_counts[remap_tf_counts >= (nrow(remap_promoters) * 0.5)] %>% names()

# what is the top TF
max(remap_tf_counts)
remap_tf_counts[which.max(remap_tf_counts)] %>% names()

zfx <- colnames(remap_promoters) %>% grep(pattern = ".ZFX.", value = TRUE)
remap_tf_counts[names(remap_tf_counts) %in% zfx] %>% graphics::boxplot()
```

```{r tf_coverage_chip_atlas}
chip_atlas_tf_counts <- chip_atlas_promoters %>%
    Matrix::colSums()

chip_atlas_tf_counts %>%
    summary()

chip_atlas_tf_counts %>%
    `/`(nrow(chip_atlas_promoters)) %>%
    summary()

chip_atlas_tf_counts %>%
    graphics::hist(breaks = 100,
                   plot = TRUE,
                   main = "ChIP-Atlas TFs coverage")

(chip_atlas_tf_counts <= 50) %>% table() / length(chip_atlas_tf_counts)

(chip_atlas_tf_counts == 0) %>% table() / length(chip_atlas_tf_counts)
(m <- chip_atlas_tf_counts == 0) %>% sum()

# How many TFs overlapp X% or more promoters
vapply(X = seq(from = 0, to = 1, by = 0.1),
       FUN = function(x) {
           sum(chip_atlas_tf_counts >= (nrow(chip_atlas_promoters) * x))
       },
       FUN.VALUE = numeric(1L)) %>%
    stats::setNames(nm = seq(from = 0, to = 100, by = 10))
# what are the top TFs?
chip_atlas_tf_counts[
    chip_atlas_tf_counts >= (nrow(chip_atlas_promoters) * 0.4)] %>%
    names()

# what is the top TF
max(chip_atlas_tf_counts)
chip_atlas_tf_counts[which.max(chip_atlas_tf_counts)] %>% names()
# "Epitope.tags_Kidney_293_SRX4892338"
# !!! this is ZfSp5a in HEK293T -- were Zf is zebrafish, this protein was
# overexpressed via transfection
```

