---
title: "Deeper look into the 'signatures'"
author: "Maciej Migdał"
output: pdf_document
params:
  promoters_ann: promoters_f5
  annotation_name: FANTOM5
---

```{r include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

devtools::load_all()
library(data.table)

promoters_ann <- get0(params$promoters_ann)

stopifnot(
  all(c(
    "SYMBOL",
    "gene_type_gencode",
    "score",
    "name",
    "tau",
    "protein_atlas_tissue_specificity",
    "encode_blacklist",
    "ensembl_sm"
  ) %in% colnames(GenomicRanges::mcols(promoters_ann)))
)
```

# Technical biases and noise vs biological signal.

Based on the Bogumil's observation that some promoters always give a signal
regardless on the transcription factor we look at. As well as, some TF gives 
signal regardless of the promoter we look at. We set out with an exploratory 
data analysis on the ReMap2020 and ChIP-Atlas data sets.

## Looking at promoter TF interaction matrix row wise (promoter wise)

ReMap2020 and ChIP-Atlas were intersected with FANTOM5 promoters into 
interaction matrices with ChIP-seq experiment as a columns and promoters as a
row names. Overall our data included `r nrow(remap_promoters)` promoters and
`r ncol(remap_promoters)`, `r ncol(chip_atlas_promoters)` ChIP-seq experiments
for ReMap2020, ChIP-Atlas respectively.

To look deeper into those data sets, first we have limited our analysis to 
core promoters defined as promoters associated with a protein coding gene (as
per GENCODE 38) and supported by ENCODE ROADMAP data. Finally, we select one
promoter per associated gene that have maximum FANTOM5 score.
   
```{r core_promoters}
# only protein coding genes as per GENCODE annotation
promoters_ann_core <- 
    promoters_ann[promoters_ann$gene_type_gencode == "protein_coding", ]

# ENCODE ROADMAP confirmation
roadmap_promoters <- rtracklayer::import.bed(
    con = system.file(
        "inst",
        "extdata",
        "Epigenome5DRoadmapDHS_promoter_hg38_liftOver.bed",
        package = "xcore"
    )
)
GenomeInfoDb::seqlevels(roadmap_promoters, pruning.mode = "coarse") <-
    GenomeInfoDb::seqlevels(promoters_ann_core)
promoters_ann_core <- intersectGR(promoters_ann_core, roadmap_promoters)

# Select best promoter per gene
best_promoters <- GenomicRanges::mcols(promoters_ann_core) %>%
    as.data.frame() %>%
    dplyr::group_by(SYMBOL) %>%
    dplyr::slice(which.max(score)) %>%
    dplyr::pull(name)
promoters_ann_core <-
    promoters_ann_core[promoters_ann_core$name %in% best_promoters, ]

# restrict interaction matrices to core promoters only
remap_promoters_core <- remap_promoters[promoters_ann_core$name, ]
chip_atlas_promoters_core <- chip_atlas_promoters[promoters_ann_core$name, ]
```

This give us an interaction matrix with `r nrow(remap_promoters_core)` promoters
(rows) and `r ncol(remap_promoters_core)`, `r ncol(chip_atlas_promoters_core)` 
ChIP-seq experiments (columns) for ReMap2020, ChIP-Atlas respectively.

### Promoter coverage

For each promoter we calculate coverage which is equal to the number
of ChIP-seq experiment the promoter intersect with.

```{r promoters_coverage}
# remap
promoters_ann_core$remap_coverage <- remap_promoters_core %>% Matrix::rowSums()
# chip_atlas
promoters_ann_core$chip_atlas_coverage <- chip_atlas_promoters_core %>% Matrix::rowSums()
```

```{r promoters_coverage_hist}
par(mfrow = c(1, 2), cex = 0.7, pty = "s")

# Promoter coverage histogram
plotHistogram(x = promoters_ann_core$remap_coverage,
             main = "ReMap2020 promoter coverage")
plotHistogram(x = promoters_ann_core$chip_atlas_coverage, 
             main = "ChIP-Atlas promoter coverage")
```

For some promoters we observe no coverage at all:
`r sum(promoters_ann_core$remap_coverage == 0)`,
`r sum(promoters_ann_core$chip_atlas_coverage == 0)`
for ReMap2020, ChIP-Atlas respectively.


### Protein Atlas gene specificity classification

We investigate promoters coverage by looking at their target genes 
classification in terms of tissue/cell type specificity as defined by Protein 
Atlas.

> The RNA specificity category is based on mRNA expression levels in a 
combination of data from HPA, GTEX and FANTOM5. The categories include: tissue enriched, group enriched, tissue enhanced, low tissue specificity and not 
detected.

>
+ Enriched - NX level in a particular tissue/region/cell type at least four 
times any other tissue/cell type.
+ Group enriched - NX levels of a group (2-5 tissues or 2-10 cell types) at 
least four times any other tissue/cell type.
+ Enhanced - NX levels of a group (1-5 tissues or 1-10 cell types) at least four 
times the mean of other tissue/cell types.
+ Low specificity - NX >= 1 in at least one tissue/cell type but not elevated in 
any tissue/cell type.
+ Not detected - NX < 1 in all tissue/cell types.

[Protein Atlas](https://www.proteinatlas.org/about/assays+annotation#classification_rna)

```{r protein_atlas_promoters}
# RNA tissue specificity
par(mfrow = c(1, 2), cex = 0.7, pty = "s")
ymax <-
  max(promoters_ann_core$remap_coverage,
      promoters_ann_core$chip_atlas_coverage)
graphics::boxplot(remap_coverage ~ protein_atlas_cell_specificity, 
                  data = promoters_ann_core,
                  ylim = c(0, ymax),
                  main = "ReMap2020",
                  ylab = "promoter coverage",
                  xlab = "",
                  las=2,
                  cex.names = 0.4)
graphics::boxplot(chip_atlas_coverage ~ protein_atlas_cell_specificity, 
                  data = promoters_ann_core,
                  ylim = c(0, ymax),
                  main = "ChIP-Atlas",
                  ylab = "promoter coverage",
                  xlab = "",
                  las=2,
                  cex.names = 0.4)
mtext(
  "Protein Atlas tissue specificity",
  side = 3,
  outer = TRUE,
  line = -2,
  font = 2,
  cex = 1.2
)
```

### Promoter tissue specificity

To generalize the idea of promoter tissue/cell type specificity to promoters 
outside Protein Atlas annotation we define specificity score based on the 
FANTOM5 CAGE data.

The  can be defined in multiple ways. Here we calculate expression based 
specificity score using $\tau$ metric defined as

$\tau = \frac{\sum_{i=1}^{n} (1-\hat{x_i})}{n-1}; \hat{x_i} = \frac{x_i}{max(x)}$

Tau varies from 0 to 1, where 0 means broadly expressed, and 1 is specific.

[reference](https://academic.oup.com/bib/article/18/2/205/2562739#119555186).

```{r promoter_f5_specificity_boxplot}
graphics::boxplot(tau ~ protein_atlas_tissue_specificity, 
                  data = promoters_ann_core,
                  main = "Tau vs Protein Atlass classififcation",
                  ylab = expression(tau),
                  xlab = "",
                  cex.axis = 0.8)
```


```{r promoters_f5_tau_hist}
plotHistogram(x = promoters_ann_core$tau,
             main = expression(paste("core promoters ", tau, " distribution")))
```

```{r promoter_f5_specificity_scatter}
par(mfrow = c(1, 2), cex = 0.7, pty = "s", mgp=c(2,1,0))
with(
  as.data.frame(promoters_ann_core),
  graphics::smoothScatter(
    x = remap_coverage,
    y = tau,
    main = sprintf(
      "ReMap2020\nSpearman: %f",
      cor(x = remap_coverage,
          y = tau,
          method = "spearman")
    ),
    ylab = expression(tau),
    xlab = "promoter coverage"
  )
)
with(
  as.data.frame(promoters_ann_core),
  graphics::smoothScatter(
    x = chip_atlas_coverage,
    y = tau,
    main = sprintf(
      "ChIP-Atlas\nSpearman: %f",
      cor(x = chip_atlas_coverage,
          y = tau,
          method = "spearman")
    ),
    ylab = expression(tau),
    xlab = "promoter coverage"
  )
)
mtext(
  "Promoter coverage vs tau",
  side = 3,
  outer = TRUE,
  line = -2,
  font = 2,
  cex = 1.2
)
```

### ENCODE blacklist

We investigate promoters coverage by overlapping the promoters with ENCODE's 
list of blacklisted regions.

>  ENCODE blacklist is a comprehensive set of regions in the human and other 
genomes that have anomalous, unstructured, or high signal in next-generation 
sequencing experiments independent of cell line or experiment. The removal of 
the ENCODE blacklist is an essential quality measure when analyzing functional 
genomics data.
[ENCODE blacklist](https://github.com/Boyle-Lab/Blacklist)

```{r blacklisted_promoters}
par(mfrow = c(1, 2), cex = 0.7, pty = "s")
ymax <-
  max(promoters_ann_core$remap_coverage,
      promoters_ann_core$chip_atlas_coverage)
with(
  as.data.frame(promoters_ann_core),
  boxplot(
    remap_coverage[encode_blacklist],
    remap_coverage[!encode_blacklist],
    names = c(sprintf(
      "blacklisted (%i)", sum(encode_blacklist)
    ),
    sprintf("other (%i)", sum(!encode_blacklist))),
    ylim = c(0, ymax),
    main = "ReMap2020",
    ylab = "promoter coverage"
  )
)
with(
  as.data.frame(promoters_ann_core),
  boxplot(
    chip_atlas_coverage[encode_blacklist],
    chip_atlas_coverage[!encode_blacklist],
    names = c(sprintf(
      "blacklisted (%i)", sum(encode_blacklist)
    ),
    sprintf("other (%i)", sum(!encode_blacklist))),
    ylim = c(0, ymax),
    main = "ChIP-Atlas",
    ylab = "promoter coverage"
  )
)
mtext(
  "ENCODE blacklisted regions",
  side = 3,
  outer = TRUE,
  line = -2,
  font = 2,
  cex = 1.2
)
```

### Promoters score (width) vs coverage

We investigate promoters coverage by looking at the relations between promoter
coverage and score, or width.

```{r promoter_score_length}
# Promoters coverage vs score
par(mfrow = c(1, 2), cex = 0.7, pty = "s")
graphics::smoothScatter(
     x = promoters_ann_core$remap_coverage,
     y = log10(promoters_ann_core$score),
     main = sprintf("ReMap2020\nSpearman: %f",
                    cor(x = promoters_ann_core$remap_coverage,
                        y = log10(promoters_ann_core$score),
                        method = "spearman")),
     xlab = "promoter coverage",
     ylab = "log10(promoter score)")
graphics::smoothScatter(
     x = promoters_ann_core$chip_atlas_coverage,
     y = log10(promoters_ann_core$score),
     main = sprintf("ChIP-Atlas\nSpearman: %f",
                    cor(x = promoters_ann_core$chip_atlas_coverage,
                        y = log10(promoters_ann_core$score),
                        method = "spearman")),
     xlab = "promoter coverage",
     ylab = "log10(promoter score)")
mtext(
  "Promoter coverage vs promoter score",
  side = 3,
  outer = TRUE,
  line = -2,
  font = 2,
  cex = 1.2
)

# Promoters coverage vs width
graphics::smoothScatter(
     x = promoters_ann_core$remap_coverage,
     y = GenomicRanges::width(promoters_ann_core),
     main = sprintf("ReMap2020\nSpearman: %f",
                    cor(x = promoters_ann_core$remap_coverage,
                        y = GenomicRanges::width(promoters_ann_core),
                        method = "spearman")),
     xlab = "promoter coverage",
     ylab = "promoter width")
graphics::smoothScatter(
     x = promoters_ann_core$chip_atlas_coverage,
     y = GenomicRanges::width(promoters_ann_core),
     main = sprintf("ChIP-Atlas\nSpearman: %f",
                    cor(x = promoters_ann_core$chip_atlas_coverage,
                        y = GenomicRanges::width(promoters_ann_core),
                        method = "spearman")),
     xlab = "promoter coverage",
     ylab = "promoter width")
mtext(
  "Promoter coverage vs promoter width",
  side = 3,
  outer = TRUE,
  line = -2,
  font = 2,
  cex = 1.2
)
```

## Looking at promoter TF interaction matrix column wise (TF wise)

### TF overlapp between ReMap2020 and ChIP-Atlas

```{r}
# ReMap2020
table((remap_meta$tf %in% chip_atlas_meta$tf)) %>%
  `/`(length(remap_meta$tf)) %>% 
  data.frame() %>% 
  dplyr::rename(TF = Var1) %>% 
  dplyr::mutate(TF = c("not included in ChIP-Atlas", "included in ChIP-Atlas")) %>%
  knitr::kable(caption = "ReMap2020 TF")

table(unique(remap_meta$tf) %in% unique(chip_atlas_meta$tf)) %>%
  `/`(length(unique(remap_meta$tf))) %>% 
  data.frame() %>% 
  dplyr::rename(TF = Var1) %>% 
  dplyr::mutate(TF = c("not included in ChIP-Atlas", "included in ChIP-Atlas")) %>%
  knitr::kable(caption = "ReMap2020 unique TF")

remap_meta$tf[! remap_meta$tf %in% chip_atlas_meta$tf] %>%
  table() %>%
  as.numeric() %>%
  plotHistogram(breaks = "Sturges", 
               main = "ReMap2020 TF not included in ChIP-Atlas")

# ChIP-Atlas
table((chip_atlas_meta$tf %in% remap_meta$tf)) %>%
  `/`(length(chip_atlas_meta$tf)) %>% 
  data.frame() %>% 
  dplyr::rename(TF = Var1) %>% 
  dplyr::mutate(TF = c("not included in ReMap2020", "included in ReMap2020")) %>%
  knitr::kable(caption = "ChIP-Atlas TF")

table(unique(chip_atlas_meta$tf) %in% unique(remap_meta$tf)) %>%
  `/`(length(unique(chip_atlas_meta$tf))) %>% 
  data.frame() %>% 
  dplyr::rename(TF = Var1) %>% 
  dplyr::mutate(TF = c("not included in ChIP-Atlas", "included in ChIP-Atlas")) %>%
  knitr::kable(caption = "ChIP-Atlas unique TF")

chip_atlas_meta$tf[! chip_atlas_meta$tf %in% remap_meta$tf] %>%
  table() %>%
  as.numeric() %>%
  plotHistogram(breaks = "Sturges", 
               main = "ChIP-Atlas TF not included in ChIP-Atlas")
```

```{r}
remap_meta[, coverage := Matrix::colSums(remap_promoters_core[, id])]
chip_atlas_meta[, coverage := Matrix::colSums(chip_atlas_promoters_core[, id])]
```

For the basic analysis we have focused only on subset of ReMap2020 and 
ChIP-Atlas. Choosing only experiments that overlap more than 1% of 
core promoters. Additionally for ReMap2020 we took only experiments without 
condition or treatment label.

Further we focused only on TF with >= 10 experiments and for ReMap2020 biotypes 
with >= 29 experiments, excluding the "breast" and "liver" biotypes, and for
ChIP-Atlas biotypes with >= 50 experiments, excluding the "Breast.cancer.cells", 
"Breast.cancer", "Prostate", "Lymphoblastoid.cell.line", "Macrophages".

```{r}
# ReMap2020
remap_meta_core <- remap_meta[(coverage >= (nrow(remap_promoters_core) * 0.01)) &
                                (condition == "")]

sel_tf <- remap_meta_core$tf %>% 
  table() %>% 
  `[`(. >= 10) %>% 
  names()
remap_meta_core <- remap_meta_core[tf %in% sel_tf]

sel_biotypes <- remap_meta_core$biotype %>% 
  table() %>% 
  `[`(. >= 29) %>% 
  names() %>%
  `[`(! . %in% c("breast", "liver"))
remap_meta_core <- remap_meta_core[biotype %in% sel_biotypes]

# annotate coverage
remap_meta_core[, cov_type := cut(
  coverage,
  breaks = c(0, quantile(coverage, 0.25), quantile(coverage, 0.75), Inf),
  labels = c("low", "medium", "high")
)]

# ChIP-Atlas
chip_atlas_meta_core <- 
  chip_atlas_meta[coverage >= (nrow(remap_promoters_core) * 0.01)]

sel_tf <- chip_atlas_meta_core$tf %>% 
  table() %>% 
  `[`(. >= 10) %>% 
  names()
chip_atlas_meta_core <- chip_atlas_meta_core[tf %in% sel_tf]

sel_biotypes <- chip_atlas_meta_core$biotype %>% 
  table() %>% 
  `[`(. >= 50) %>% 
  names() %>%
  `[`(! . %in% c("Breast.cancer.cells", "Breast.cancer", "Prostate", "Lymphoblastoid.cell.line", "Macrophages"))
chip_atlas_meta_core <- chip_atlas_meta_core[biotype %in% sel_biotypes]

# annotate coverage
chip_atlas_meta_core[, cov_type := cut(
  coverage,
  breaks = c(0, quantile(coverage, 0.25), quantile(coverage, 0.75), Inf),
  labels = c("low", "medium", "high")
)]
```

```{r}
par(mfrow = c(1, 2), cex = 0.7, pty = "s")
plotHistogram(remap_meta_core$coverage, 
              main = "ReMap2020")
plotHistogram(chip_atlas_meta_core$coverage, 
              main = "ChIP-Atlas")
mtext(
  "Selected experiment's coverage",
  side = 3,
  outer = TRUE,
  line = -2,
  font = 2,
  cex = 1.2
)
```

### TF / biotype confusion matrix

```{r tf_bg_confusion_matrix_remap, echo=FALSE}
remap_tfs_tab <-
  data.table::dcast(data = remap_meta_core,
                    formula = biotype ~ tf,
                    fun = length)
ord <- c(1, order(remap_tfs_tab[, -1] %>% colSums(), decreasing = TRUE) + 1)
data.table::setcolorder(remap_tfs_tab, ord)
remap_tfs_tab[, .r := -rowSums(remap_tfs_tab[, -1])]
data.table::setorder(remap_tfs_tab, .r)[, .r := NULL]

knitr::kable(remap_tfs_tab[, 1:11], caption = "ReMap TF/biotype matrix, top 10 TF")
```

```{r tf_bg_confusion_matrix_chip_atlas, echo=FALSE}
chip_atlas_tfs_tab <-
  data.table::dcast(data = chip_atlas_meta_core,
                    formula = biotype ~ tf,
                    fun = length)
ord <- c(1, order(chip_atlas_tfs_tab[, -1] %>% colSums(), decreasing = TRUE) + 1)
data.table::setcolorder(chip_atlas_tfs_tab, ord)
chip_atlas_tfs_tab[, .r := -rowSums(chip_atlas_tfs_tab[, -1])]
data.table::setorder(chip_atlas_tfs_tab, .r)[, .r := NULL]

knitr::kable(chip_atlas_tfs_tab[, 1:11], caption = "ChIP-Atlas TF/biotype matrix, top 10 TF")
```

## Choosing distance measure
In order to select a distance measure that would best describe associations
between experiments, we have looked into several distance measures: Phi 
(which we approximate with Pearson's), Gower, Jaccard, Dice, Yule, Russel, 
Tanimoto. In order to select one we wanted to maximize the difference between
mean distance between the same TF and different TF.

The selection was done using core ReMap2020.

Outliers were not marked on plots.

```{r select_distance_measure}
mat <- remap_promoters_core[, remap_meta_core[, id]] %>% as.matrix()

getDists <- function(mat, patterns, invert=FALSE) {
  sapply(
    X = patterns, 
    function(x) {
      m_row <- grep(x, rownames(mat), value = TRUE)
      m_col <- grep(x, colnames(mat), value = TRUE, invert = invert)
      sel_cor_mat <- mat[m_row, m_col]
      sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
    })
}

compDists <- function(D, Dnames, patterns) {
  Ds <- lapply(
    X = seq_along(D),
    FUN = function(i) {
      same_tf <- data.frame(metric = Dnames[i], 
                            val = unlist(getDists(D[[i]], patterns)),
                            type = "same")                
      diff_tf <- data.frame(metric = Dnames[i],
                            val = unlist(getDists(D[[i]], patterns, invert=TRUE)),
                            type = "diff.")
      rbind(same_tf, diff_tf)
    }
  ) %>% do.call(what = rbind)
}

# pearson approximates phi coefficient
d_pearson <- (1 - cor(mat, method = "pearson")) / 2
# d_binary <- dist(t(mat), method = "binary") %>% as.matrix() equivalent to jaccard
d_gower <- cluster::daisy(t(mat), metric = "gower", warnType = FALSE) %>% as.matrix()
d_jaccard <- proxy::dist(x = mat, 
                         method = "Jaccard", 
                         by_rows = FALSE, 
                         convert_similarities = TRUE) %>% as.matrix()
d_dice <- proxy::dist(x = mat, 
                         method = "Dice", 
                         by_rows = FALSE, 
                         convert_similarities = TRUE) %>% as.matrix()
# here it gives distance but it actuall varies from -1 to + 1 so I could plot it on heatmap
d_yule <- proxy::dist(x = mat, 
                      method = "Yule", 
                      by_rows = FALSE, 
                      convert_similarities = TRUE) %>% as.matrix()
d_russel <- proxy::dist(x = mat, 
                         method = "Russel", 
                         by_rows = FALSE, 
                         convert_similarities = TRUE) %>% as.matrix()
d_tanimoto <- proxy::dist(x = mat, 
                         method = "Tanimoto", 
                         by_rows = FALSE, 
                         convert_similarities = TRUE) %>% as.matrix()

d_comparison <- compDists(
  list(d_pearson, d_gower, d_jaccard, d_dice, d_yule, d_russel, d_tanimoto), 
  c("Phi", "Gower", "Jaccard", "Dice", "Yule", "Russel", "Tanimoto"), 
  remap_meta_core[, unique(tf)])
par(mar = c(10, 4, 4, 1))
boxplot(val ~ type + metric, 
        data = d_comparison,
        sep = " ", 
        las=2,
        ylim = c(0, 1),
        outline = FALSE,
        main = "Distances between experiments grouped by TF",
        ylab = "distance",
        xlab = "")

d_comparison %>%
  dplyr::group_by(type, metric) %>%
  dplyr::summarise(mean = mean(val)) %>%
  dplyr::group_by(metric) %>%
  dplyr::summarise(dAve = abs(mean[1] - mean[2])) %>%
  knitr::kable(col.names = c("Distance metric", "$\\Delta Average$"), 
               caption = "Difference between mean distance between same and different TF groups",
               escape = FALSE)
```

Explore Yule's coefficient of colligation

```{r}
par(mfrow = c(1, 3), cex = 0.7, pty = "s")

# tf
same_tf_cor <- getDists(d_yule, remap_meta_core[, unique(tf)]) %>% 
  unlist()
diff_tf_cor <- getDists(d_yule, remap_meta_core[, unique(tf)], invert = TRUE) %>% 
  unlist()
boxplot(
  same_tf_cor, 
  diff_tf_cor, 
  names = c(sprintf("same (%i)", length(same_tf_cor)),
            sprintf("different (%i)", length(diff_tf_cor))),
  ylim = c(0, 1),
  outline = FALSE,
  main = "Grouped by TF",
  ylab = "Yule's distance")

# biotype
same_bio_cor <- getDists(d_yule, remap_meta_core[, unique(biotype)]) %>%
  unlist()
diff_bio_cor <- getDists(d_yule, remap_meta_core[, unique(biotype)], invert = TRUE) %>%
  unlist()
boxplot(
  same_bio_cor, diff_bio_cor, 
  names = c(sprintf("same (%i)", length(same_bio_cor)),
            sprintf("different (%i)", length(diff_bio_cor))),
  ylim = c(0, 1),
  outline = FALSE,
  main = "Grouped by biotype",
  ylab = "Yule's distance")

# study
# TODO make it work on metadata
rownames(d_yule) <- paste0(rownames(d_yule), "=", remap_meta_core[, study])
colnames(d_yule) <- paste0(colnames(d_yule), "=", remap_meta_core[, study]) 
same_study_cor <- getDists(d_yule, remap_meta_core[, unique(study)]) %>%
  unlist()
diff_sutdy_cor <- getDists(d_yule, remap_meta_core[, unique(study)], invert = TRUE) %>%
  unlist()
rownames(d_yule) <- sub("=.*", "", rownames(d_yule))
colnames(d_yule) <- sub("=.*", "", colnames(d_yule))

boxplot(
  same_study_cor, diff_sutdy_cor, 
  names = c(sprintf("same (%i)", length(same_study_cor)),
            sprintf("different (%i)", length(diff_sutdy_cor))),
  ylim = c(0, 1),
  outline = FALSE,
  main = "Grouped by study",
  ylab = "Yule's distance")

mtext(
  "Yule's distance between experiments",
  side = 3,
  outer = TRUE,
  line = -2,
  font = 2,
  cex = 1.2)
```

### MYC

```{r}
breaks <- seq(from = -1, to = 1, by = 0.1)
color <-
  grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(
    n = 7, 
    name = "RdYlBu"))
  )(length(breaks))
  
m <- remap_meta_core[tf == "MYC", id]
mat <- d_yule[m, m]
plotExperimentsHeatmap(
  mat = 1 - (2 * mat),
  meta = remap_meta_core[tf == "MYC", -c("tf", "condition", "tf_dbd")],
  clustering_distance_rows = as.dist(mat),
  clustering_distance_cols = as.dist(mat),
  clustering_method = "ward.D2",
  breaks = breaks,
  color = color,
  labels_row = remap_meta_core[tf == "MYC", biotype],
  labels_col = remap_meta_core[tf == "MYC", biotype],
  main = "MYC experiments Yule's Y coefficient")
```

### MAX

```{r}
breaks <- seq(from = -1, to = 1, by = 0.1)
color <-
  grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(
    n = 7, 
    name = "RdYlBu"))
  )(length(breaks))

m <- remap_meta_core[tf == "MAX", id]
mat <- d_yule[m, m]
plotExperimentsHeatmap(
  mat =  1 - (2 * mat),
  meta = remap_meta_core[tf == "MAX", -c("tf", "condition", "tf_dbd")],
  clustering_distance_rows = as.dist(mat),
  clustering_distance_cols = as.dist(mat),
  clustering_method = "ward.D2",
  breaks = breaks,
  color = color,
  labels_row = remap_meta_core[tf == "MAX", biotype],
  labels_col = remap_meta_core[tf == "MAX", biotype],
  main = "MAX experiments Yule's Y coefficient")
```

### HDAC2

```{r}
breaks <- seq(from = -1, to = 1, by = 0.1)
color <-
  grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(
    n = 7, 
    name = "RdYlBu"))
  )(length(breaks))

m <- remap_meta_core[tf == "HDAC2", id]
mat <- d_yule[m, m]
plotExperimentsHeatmap(
  mat = 1 - (2 * mat),
  meta = remap_meta_core[tf == "HDAC2", -c("tf", "condition", "tf_dbd")],
  clustering_distance_rows = as.dist(mat),
  clustering_distance_cols = as.dist(mat),
  clustering_method = "ward.D2",
  breaks = breaks,
  color = color,
  labels_row = remap_meta_core[tf == "HDAC2", biotype],
  labels_col = remap_meta_core[tf == "HDAC2", biotype],
  main = "HDAC2 experiments Yule's Y coefficient")
```

### FOXA1

```{r}
breaks <- seq(from = -1, to = 1, by = 0.1)
color <-
  grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(
    n = 7, 
    name = "RdYlBu"))
  )(length(breaks))

m <- remap_meta_core[tf == "FOXA1", id]
mat <- d_yule[m, m]
plotExperimentsHeatmap(
  mat = 1 - (2 * mat),
  meta = remap_meta_core[tf == "FOXA1", -c("tf", "condition", "tf_dbd")],
  clustering_distance_rows = as.dist(mat),
  clustering_distance_cols = as.dist(mat),
  clustering_method = "ward.D2",
  breaks = breaks,
  color = color,
  labels_row = remap_meta_core[tf == "FOXA1", biotype],
  labels_col = remap_meta_core[tf == "FOXA1", biotype],
  main = "FOXA1 experiments Yule's Y association")
```

### REST

```{r}
breaks <- seq(from = -1, to = 1, by = 0.1)
color <-
  grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(
    n = 7, 
    name = "RdYlBu"))
  )(length(breaks))

m <- remap_meta_core[tf == "REST", id]
mat <- d_yule[m, m]
plotExperimentsHeatmap(
  mat = 1 - (2 * mat),
  meta = remap_meta_core[tf == "REST", -c("tf", "condition", "tf_dbd")],
  clustering_distance_rows = as.dist(mat),
  clustering_distance_cols = as.dist(mat),
  clustering_method = "ward.D2",
  breaks = breaks,
  color = color,
  labels_row = remap_meta_core[tf == "REST", biotype],
  labels_col = remap_meta_core[tf == "REST", biotype],
  main = "REST experiments Yule's Y association")
```

### Core experiments heatmap

```{r include=FALSE, eval=FALSE}
breaks <- seq(from = -1, to = 1, by = 0.1)
color <-
  grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(
    n = 7, 
    name = "RdYlBu"))
  )(length(breaks))

plotExperimentsHeatmap(
  mat = 1 - (2 * d_yule),
  meta = remap_meta_core[, -c("study", "condition")],
  clustering_distance_rows = as.dist(d_yule),
  clustering_distance_cols = as.dist(d_yule),
  clustering_method = "ward.D2",
  breaks = breaks,
  color = color,
  labels_row = remap_meta_core[, paste(tf, biotype)],
  labels_col = remap_meta_core[, paste(tf, biotype)],
  main = "Core experiments Yule's Y association")
```

## Yule's distance bulk calculation

```{r eval=FALSE}
# exclude experiments with coverage below 1% of core promoters

j <- Matrix::colSums(remap_promoters_core) >= (length(promoters_ann_core) * 0.01)
d_yule_remap <- proxy::dist(x = as.matrix(remap_promoters_core[, j]), 
                         method = "Yule", 
                         by_rows = FALSE, 
                         convert_similarities = TRUE)
save(d_yule_remap, file = "d_yule_remap.rda")

j <- Matrix::colSums(chip_atlas_promoters_core) >= (length(promoters_ann_core) * 0.01)
d_yule_chip_atlas <- proxy::dist(x = as.matrix(chip_atlas_promoters_core[, j]), 
                         method = "Yule", 
                         by_rows = FALSE, 
                         convert_similarities = TRUE)
save(d_yule_chip_atlas, file = "d_yule_chip_atlas.rda")
```

```{r}
load(system.file("inst", "analysis", "d_yule_remap.rda", package = "xcore"))
load(system.file("inst", "analysis", "d_yule_chip_atlas.rda", package = "xcore"))
```

## Clustering experiments

### Hierarchical Clustering

```{r}
measureClust <- function(c1, hc, N) {
  nclust_measure <- lapply(
    X = N,
    FUN = function(n) {
      do.call(cbind,
              aricode::clustComp(c1 = c1,
                                 c2 = cutree(hc, n)))
    }
  )
  do.call(rbind, nclust_measure)
}

plotMeasureClust <- function(n_cl, 
                             cl_measure, 
                             measure,
                             ylim,
                             pos = "topright",
                             main = "",
                             add_legend = TRUE) {
  cols <- colorspace::qualitative_hcl(length(cl_measure))
  plot(
    x = NA,
    y = NA,
    main = main,
    xlim = c(min(n_cl), max(n_cl)),
    ylim = ylim,
    xlab = "number of clusters",
    ylab = measure
  )
  for (i in seq_along(cl_measure)) {
    points(n_cl, cl_measure[[i]][, measure], col = cols[i], pch = i)
  }
  if (add_legend) {
    legend(
      pos,
      legend = names(cl_measure),
      col = cols,
      pch = seq_along(cl_measure))
  }
}
```

### Linkage method choice

```{r}
nclust <- seq(from = 1, to = 400, by = 5)
linkage_methods <- c("complete", "single", "ward.D", "ward.D2", "average", 
                     "mcquitty", "median", "centroid")

# ReMap2020
id_core <- remap_meta_core[, id]
d_core <- as.dist(as.matrix(d_yule_remap)[id_core, id_core])
true_cl <- remap_meta_core[, tf]
clust_measures <- lapply(
  X = linkage_methods, 
  FUN = function(m) {
    hc <- hclust(d_core, method = m)
    measureClust(true_cl, hc, nclust)
  })
names(clust_measures) <- linkage_methods

par(mfrow = c(1, 2), oma = c(0, 0, 4, 0), cex = 0.7, pty = "s")
plotMeasureClust(
  nclust,
  clust_measures,
  "ARI",
  ylim = c(0, 0.65),
  pos = "topright",
  main = "Adjusted Rand Index",
  add_legend = FALSE
)
plotMeasureClust(
  nclust,
  clust_measures,
  "AMI",
  ylim = c(0, 0.8),
  pos = "bottomright",
  main = "Adjusted Mutual Information"
)
mtext(
  "Hierarchical clustering performance\nfor different linkage methods\nReMap2020 core experiments",
  side = 3,
  outer = TRUE,
  line = -1,
  font = 2,
  cex = 0.9
)

# ChIP-Atlas
id_core <- chip_atlas_meta_core[, id]
d_core <- as.dist(as.matrix(d_yule_chip_atlas)[id_core, id_core])
true_cl <- chip_atlas_meta_core[, tf]
clust_measures <- lapply(
  X = linkage_methods, 
  FUN = function(m) {
    hc <- hclust(d_core, method = m)
    measureClust(true_cl, hc, nclust)
  })
names(clust_measures) <- linkage_methods

plotMeasureClust(
  nclust,
  clust_measures,
  "ARI",
  ylim = c(0, 0.65),
  pos = "topright",
  main = "Adjusted Rand Index"
)
plotMeasureClust(
  nclust,
  clust_measures,
  "AMI",
  ylim = c(0, 0.8),
  pos = "bottomright",
  main = "Adjusted Mutual Information"
)
mtext(
  "Hierarchical clustering performance\nfor different linkage methods\nChIP-Atlas core experiments",
  side = 3,
  outer = TRUE,
  line = -1,
  font = 2,
  cex = 0.9
)
```

### clustering as compared to known labels

```{r}
nclust <- seq(from = 1, to = 400, by = 5)
labels <- c("tf", "biotype", "study")

# ReMap2020
id_core <- remap_meta_core[, id]
d_core <- as.dist(as.matrix(d_yule_remap)[id_core, id_core])
labs_measure <- lapply(
  X = labels, 
  FUN = function(l) {
    hc <- hclust(d_core, method = "ward.D2")
    measureClust(remap_meta_core[[l]], hc, nclust)
  })
names(labs_measure) <- labels

par(mfrow = c(1, 2), oma = c(0, 0, 3, 0), cex = 0.7, pty = "s")
plotMeasureClust(
  nclust,
  labs_measure,
  "ARI",
  ylim = c(0, 0.65),
  pos = "topright",
  main = "Adjusted Rand Index"
)
plotMeasureClust(
  nclust,
  labs_measure,
  "AMI",
  ylim = c(0, 0.8),
  pos = "bottomright",
  main = "Adjusted Mutual Information"
)
mtext(
  "Hierarchical clustering performance for different labels\nReMap2020 selected experiments",
  side = 3,
  outer = TRUE,
  line = -1,
  font = 2,
  cex = 1
)

# ChIP-Atlas
id_core <- chip_atlas_meta_core[, id]
d_core <- as.dist(as.matrix(d_yule_chip_atlas)[id_core, id_core])
labs_measure <- lapply(
  X = labels, 
  FUN = function(l) {
    hc <- hclust(d_core, method = "ward.D2")
    measureClust(chip_atlas_meta_core[[l]], hc, nclust)
  })
names(labs_measure) <- labels

plotMeasureClust(
  nclust,
  labs_measure,
  "ARI",
  ylim = c(0, 0.65),
  pos = "topright",
  main = "Adjusted Rand Index"
)
plotMeasureClust(
  nclust,
  labs_measure,
  "AMI",
  ylim = c(0, 0.8),
  pos = "bottomright",
  main = "Adjusted Mutual Information"
)
mtext(
  "Hierarchical clustering performance for different labels\nChIP-Atlas selected experiments",
  side = 3,
  outer = TRUE,
  line = -1,
  font = 2,
  cex = 1
)
```

### Partitioning Around Medoids

```{r}
nclust <- seq(from = 1, to = 400, by = 10)
labels <- c("tf", "biotype", "study")

# ReMap2020
id_core <- remap_meta_core[, id]
d_core <- as.dist(as.matrix(d_yule_remap)[id_core, id_core])

cl_pam <- lapply(
  X = nclust, 
  FUN = function(i) {
    cluster::pam(d_core, k = i, cluster.only = TRUE)
  })
cl_measure <- lapply(
  X = labels, 
  FUN = function(l) {
    cbind(
      ARI = sapply(cl_pam, aricode::ARI, c1 = remap_meta_core[[l]]),
      AMI = sapply(cl_pam, aricode::AMI, c1 = remap_meta_core[[l]])
    )
  })
names(cl_measure) <- labels

par(mfrow = c(1, 2), oma = c(0, 0, 4, 0), cex = 0.7, pty = "s")
plotMeasureClust(
  nclust,
  cl_measure,
  "ARI",
  ylim = c(0, 0.65),
  pos = "topright",
  main = "Adjusted Rand Index"
)
plotMeasureClust(
  nclust,
  cl_measure,
  "AMI",
  ylim = c(0, 0.8),
  pos = "bottomright",
  main = "Adjusted Mutual Information"
)
mtext(
  "Partitioning Around Medoids clustering performance\nfor different labels\nReMap2020 selected experiments",
  side = 3,
  outer = TRUE,
  line = -1,
  font = 2,
  cex = 1.2
)

# ChIP-Atlas
# id_core <- chip_atlas_meta_core[, id]
# d_core <- as.dist(as.matrix(d_yule_chip_atlas)[id_core, id_core])
# 
# cl_pam <- lapply(
#   X = nclust,
#   FUN = function(i) {
#     cluster::pam(d_core, k = i, cluster.only = TRUE)
#   })
# cl_measure <- lapply(
#   X = labels,
#   FUN = function(l) {
#     cbind(
#       ARI = sapply(cl_pam, aricode::ARI, c1 = chip_atlas_meta_core[[l]]),
#       AMI = sapply(cl_pam, aricode::AMI, c1 = chip_atlas_meta_core[[l]])
#     )
#   })
# names(cl_measure) <- labels
# 
# plotMeasureClust(
#   nclust,
#   cl_measure,
#   "ARI",
#   ylim = c(0, 0.65),
#   pos = "topright",
#   main = "Adjusted Rand Index"
# )
# plotMeasureClust(
#   nclust,
#   cl_measure,
#   "AMI",
#   ylim = c(0, 0.8),
#   pos = "bottomright",
#   main = "Adjusted Mutual Information"
# )
# mtext(
#   "Partitioning Around Medoids clustering performance\nfor different labels\nChIP-Atlas selected experiments",
#   side = 3,
#   outer = TRUE,
#   line = -1,
#   font = 2,
#   cex = 1.2
# )
```

## Experiment collapsing

```{r eval=FALSE}
# ReMap2020
ids_remap <- labels(d_yule_remap)
meta_remap <- remap_meta[id %in% ids_remap, ]
hc_remap <- hclust(d_yule_remap, method = "ward.D2")
 
# dynamic cut parameters selection
res <- selectParams4dynamicTreeCut(
  dendro = hc_remap, 
  distM = as.matrix(d_yule_remap), 
  ref_cl = meta_remap[["tf"]])
res[res[, (AMI * ARI) == max(AMI * ARI)]]
#    deepSplit cutHeight       ARI      AMI   N
# 1:         1       2.5 0.4389195 0.730026 374

cl_remap <- dynamicTreeCut::cutreeDynamic(
  dendro = hc_remap,
  cutHeight = 2.5,
  minClusterSize = 3,
  method = "hybrid",
  distM = as.matrix(d_yule_remap),
  deepSplit = 1,
  pamStage = TRUE)

remap_promoters_core_collapsed <- 
  collapseInteractionMatrix(mat = remap_promoters_core[, ids_remap], 
                            meta = meta_remap, 
                            cl = cl_remap, 
                            alpha = 0.5, # 0.5 corresponds to majority vote
                            purity_feature = "tf", 
                            min_pruning_purity = 0.5) # clusters with purity > 0.5 are pruned
save(remap_promoters_core_collapsed, file = "remap_promoters_core_collapsed.rda")

remap_promoters_collapsed <- 
  collapseInteractionMatrix(mat = remap_promoters[, ids_remap], 
                            meta = meta_remap, 
                            cl = cl_remap, 
                            alpha = 0.5, # 0.5 corresponds to majority vote
                            purity_feature = "tf", 
                            min_pruning_purity = 0.5) # clusters with purity > 0.5 are pruned
save(remap_promoters_core_collapsed, file = "remap_promoters_collapsed.rda")

meta_remap[, cluster := cl_remap]
data.table::fwrite(meta_remap, file = "remap_metadata_clusters.csv")

# ChIP-Atlas
ids_chip_atlas <- labels(d_yule_chip_atlas)
meta_chip_atlas <- chip_atlas_meta[id %in% ids_chip_atlas, ]
hc_chip_atlas <- hclust(d_yule_chip_atlas, method = "ward.D2")
 
# dynamic cut parameters selection
res <- selectParams4dynamicTreeCut(
  dendro = hc_chip_atlas, 
  distM = as.matrix(d_yule_chip_atlas), 
  ref_cl = meta_chip_atlas[["tf"]])
res[res[, (AMI * ARI) == max(AMI * ARI)]]
#    deepSplit cutHeight       ARI       AMI   N
# 1:         2       5.7 0.5661968 0.7048017 377

cl_chip_atlas <- dynamicTreeCut::cutreeDynamic(
  dendro = hc_chip_atlas,
  cutHeight = 5.7,
  minClusterSize = 3,
  method = "hybrid",
  distM = as.matrix(d_yule_chip_atlas),
  deepSplit = 2,
  pamStage = TRUE)

chip_atlas_promoters_core_collapsed <- 
  collapseInteractionMatrix(mat = chip_atlas_promoters_core[, ids_chip_atlas], 
                            meta = meta_chip_atlas, 
                            cl = cl_chip_atlas, 
                            alpha = 0.5, # 0.5 corresponds to majority vote
                            purity_feature = "tf", 
                            min_pruning_purity = 0.5) # clusters with purity > 0.5 are pruned
save(chip_atlas_promoters_core_collapsed, file = "chip_atlas_promoters_core_collapsed.rda")

chip_atlas_promoters_collapsed <- 
  collapseInteractionMatrix(mat = chip_atlas_promoters[, ids_chip_atlas], 
                            meta = meta_chip_atlas, 
                            cl = cl_chip_atlas, 
                            alpha = 0.5, # 0.5 corresponds to majority vote
                            purity_feature = "tf", 
                            min_pruning_purity = 0.5) # clusters with purity > 0.5 are pruned
save(chip_atlas_promoters_core_collapsed, file = "chip_atlas_promoters_collapsed.rda")

meta_chip_atlas[, cluster := cl_chip_atlas]
data.table::fwrite(meta_chip_atlas, file = "chip_atlas_metadata_clusters.csv")
```

# Explore clustering and collapsed matrices

```{r}
meta_remap <- data.table::fread("remap_metadata_clusters.csv")
meta_chip_atlas <- data.table::fread("chip_atlas_metadata_clusters.csv")

remap_ncl <- meta_remap[, length(unique(cluster))]
chip_atlas_ncl <- meta_chip_atlas[, length(unique(cluster))]

par(mfrow = c(1, 2), oma = c(0, 0, 0, 0), cex = 0.7, pty = "s")
plotHistogram(
  x = meta_remap[, .N, by = cluster][["N"]], 
  breaks = "scott", 
  main = sprintf("ReMap2020 (# clusters: %i)", remap_ncl))
plotHistogram(
  x = meta_chip_atlas[, .N, by = cluster][["N"]], 
  breaks = "scott", 
  main = sprintf("ChIP-Atlas (# clusters: %i)", chip_atlas_ncl))
mtext(
  "Clusters size distribution",
  side = 3,
  outer = TRUE,
  line = -4,
  font = 2,
  cex = 1.2
)

# purity_dt <- meta_remap[, .(
#   tf_purity = vectorPurity(tf),
#   biotype_purity = vectorPurity(biotype),
#   study_purity = vectorPurity(study),
#   dbd_purity = vectorPurity(tf_dbd),
#   .N
#   ),
# by = cluster]
# 
# par(mfrow = c(1, 1), oma = c(0, 0, 0, 0), cex = 0.7, pty = "s")
# 
# # clustered by study?
# plotSubtree(hc_remap, meta_remap[cluster == 92, ], "tf")
# mtext("Cluster 92, clustering by study?")
# plotSubtree(hc_remap, meta_remap[cluster == 175, ], "tf")
# mtext("Cluster 175, clustering by study?")
# 
# # clustered by biotype?
# plotSubtree(hc_remap, meta_remap[cluster == 46, ], "tf")
# mtext("Cluster 46, clustering by biotye?")
# plotSubtree(hc_remap, meta_remap[cluster == 110, ], "tf")
# mtext("Cluster 110, clustering by biotype?")
```
