---
title: "Deeper look into the 'signatures'"
author: "Maciej Migdał"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

```{r include=FALSE}
knitr::opts_chunk$set(warning = FALSE)

devtools::load_all()
library(data.table)
```

```{r functions_data, cache=TRUE, echo=FALSE, message=FALSE}
# Internally used functions

# Function return logical vector indicating columns populated in all rows
# mat - object of class dgCMatrix with only 1 and 0 values
getSharedColumns <- function(mat) {
    stopifnot(is(mat, "dgCMatrix"))
    stopifnot(nrow(mat) > 0)

    Matrix::colSums(mat) == nrow(mat)
}

# Funtion return features from subject overlapped by query
# where subject, query are GRanges objects
# ... other arguments internally passed to GenomicRanges::findOverlaps
intersectGr <- function(subject, query, drop.duplicates = TRUE, ...) {
    stopifnot(is(subject, "GRanges"))
    stopifnot(is(query, "GRanges"))

    hits <- GenomicRanges::findOverlaps(query = query, subject = subject, ...)
    i <- S4Vectors::subjectHits(hits)
    
    if (drop.duplicates) {
      i <- i[! duplicated(i)]
    }
    
    subject[i, ]
}

# Function returns all pairs of values in a vector; order in pair is not
# considered relevant
# x character vector of length > 1
getSharedColumnsPairs <- function(mat) {
    stopifnot(is(mat, "dgCMatrix"))
    stopifnot(nrow(mat) > 0)
    
    indexes <- combn(x = seq(from = 1, to = nrow(mat)), 
                     m = 2,
                     simplify = FALSE)
    vapply(X = indexes, 
           function(i) sum(Matrix::colSums(mat[i, ]) == 2), 
           FUN.VALUE = integer(1L))
}

# Function return summary as a string
summaryString <- function(...) {
    s <- summary(...)
    snm <- names(s)
    sval <- s %>% round(2) %>% as.character()
    len <- vapply(X = seq_along(snm), 
                  FUN = function(i) max(nchar(c(snm[i], sval[i]))) + 3L, 
                  FUN.VALUE = integer(1L))
    snm <- paste(vapply(
        X = seq_along(snm),
        FUN = function(i) stringr::str_pad(snm[i], len[i]),
        FUN.VALUE = character(1L)
        ), collapse = " ")
    sval <- paste(vapply(
        X = seq_along(sval),
        FUN = function(i) stringr::str_pad(sval[i], len[i] + 2),
        FUN.VALUE = character(1L)
        ), collapse = " ")
    paste0(snm, "\n", sval)
}

# Function calculates TF or Context coverage based on interaction matrix
# col to use for grouping
getCoverage <- function(mat, col = c(2, 3), sep = "\\.") {
    n <- stringr::str_count(colnames(mat)[1], sep)
    colnm <- colnames(mat) %>% 
        gsub(pattern = paste(c("(.*)", rep(c(sep, "(.*)"), n)), collapse = ""),
             replacement = paste0("\\", col))
    DelayedArray::colsum(x = mat, group = colnm %>% as.factor()) %>%
        `>`(1) %>%
        rowSums()
}

# Function plot coverage histogram
coverageHist<- function(x, breaks = 40, cex = 0.7, xlab = "", ...) {
    par(mar = c(5, 5, 6, 2))
    graphics::hist(x = x, 
                   breaks = breaks,
                   cex = 0.7,
                   xlab = xlab,
                   ...)
    mtext(text = summaryString(x), 
          side = 3,
          cex=0.6)
}

# Function plot two group box plot
blacklistBoxPlot <- function(coverage, cov_names, gr1, aname, bname, ...) {
    blk <- coverage[cov_names %in% gr1]
    nonblk <- coverage[! cov_names %in% gr1]
    aname <- paste0(aname, " (%i)")
    bname <- paste0(bname, " (%i)")

    graphics::boxplot(blk, nonblk,
                      names = c(sprintf(aname, length(blk)), 
                                sprintf(bname, length(nonblk))),
                      ...)
}


# tau calculation from on https://academic.oup.com/bib/article/18/2/205/2562739
#Function require a vector with expression of one gene in different tissues.
#If expression for one tissue is not known, gene specificity for this gene is NA
#Minimum 2 tissues
fTau <- function(x)
{
	if(all(!is.na(x)))
 	{
 		if(min(x, na.rm=TRUE) >= 0)
		{
 			if(max(x)!=0)
 			{
 				x <- (1-(x/max(x)))
 				res <- sum(x, na.rm=TRUE)
 				res <- res/(length(x)-1)
 			} else {
 				res <- 0
 			}
 		} else {
 		res <- NA
 		#print("Expression values have to be positive!")
 		} 
 	} else {
 		res <- NA
 		#print("No data for this gene avalable.")
 	} 
 	return(res)
}

# Internally used datasets
# ENCODE blacklist list 636 regions spanning ~7% of human genome
blacklist <- rtracklayer::import(system.file("inst",
                                             "extdata",
                                             "hg38-blacklist.v2.bed.gz",
                                             package = "xcore")) %>%
    intersectGr(subject = c(xcore::promoters_f5, xcore::enhancers), 
                type = "any") %>%
    GenomicRanges::mcols() %>%
    `[[`("name")
    

# ENSEMBL soft mask list 4725854 regions spanning ~50% of human genome
ensembl_sm <- rtracklayer::import(system.file("inst",
                                              "extdata",
                                              "hg38_ensembl_sm.bed.gz",
                                              package = "xcore")) %>%
    intersectGr(subject = c(xcore::promoters_f5, xcore::enhancers), 
                type = "any") %>%
    GenomicRanges::mcols() %>%
    `[[`("name")

# Protein atlas
protein_atlas <-
    data.table::fread(
        file = system.file("inst", "extdata", "proteinatlas.tsv.gz", package = "xcore"),
        header = TRUE
    )[! duplicated(Gene), ]

# FANTOM5 promoter expression table
# Matrix of promoters normalized expression across datasets in FANTOM5, 
# column names are not imported!
promoters_f5_expression <-
    data.table::fread(
        file = system.file("inst",
                           "extdata",
                           "hg38_fair+new_CAGE_peaks_phase1and2_tpm.osc.txt.gz",
                           package = "xcore"),
        header = TRUE
    )

promoters_f5_expression <-
    data.table::melt(data = promoters_f5_expression,
                     id.vars = "00Annotation")
promoters_f5_expression <- promoters_f5_expression[, .(
                         mean = mean(value, na.rm = TRUE),
                         min = min(value, na.rm = TRUE),
                         max = max(value, na.rm = TRUE),
                         sd = sd(value, na.rm = TRUE),
                         tau = fTau(log(value + 1))
                     ),
                     by = "00Annotation"]

# enhancers expression
enhancers_f5_expression <-
    data.table::fread(
        file = system.file("inst",
                           "extdata",
                           "F5.hg38.enhancers.expression.tpm.matrix.gz",
                           package = "xcore"),
        header = TRUE
    )

enhancers_f5_expression <-
    data.table::melt(data = enhancers_f5_expression,
                     id.vars = "V1")
enhancers_f5_expression <- enhancers_f5_expression[, .(
                         mean = mean(value, na.rm = TRUE),
                         min = min(value, na.rm = TRUE),
                         max = max(value, na.rm = TRUE),
                         sd = sd(value, na.rm = TRUE),
                         tau = fTau(log(value + 1))
                     ),
                     by = "V1"]

gc()
```
# Technical biases and noise vs biological signal.

Based on the Bogumil's observation that some promoters always give a signal
regardless on the transcription factor we look at. As well as, some TF gives 
signal regardless of the promoter we look at. We set out with an exploratory 
data analysis on the ReMap2020 and ChIP-Atlas data sets.

## Looking at promoter TF interaction matrix row wise (promoter wise)

ReMap2020 and ChIP-Atlas were intersected with FANTOM5 promoters into 
interaction matrices with ChIP-seq experiment as a columns and promoters as a
row names. Overall our data included `r nrow(remap_promoters)` promoters and
`r ncol(remap_promoters)`, `r ncol(chip_atlas_promoters)` ChIP-seq experiments
for ReMap2020, ChIP-Atlas respectively.

To look deeper into those data sets, first we have limited our analysis to 
core promoters defined as promoters associated with a protein coding gene (as
per GENCODE 38) and supported by ENCODE ROADMAP data. Finally, we select one
promoter per associated gene that have maximum FANTOM5 score.
   
```{r core_promoters, echo=FALSE}
# only protein coding genes as per GENCODE annotation
promoters_f5_core <- 
    promoters_f5[promoters_f5$gene_type_gencode == "protein_coding", ]

# ENCODE ROADMAP confirmation
roadmap_promoters <- rtracklayer::import.bed(
    con = system.file(
        "inst",
        "extdata",
        "Epigenome5DRoadmapDHS_promoter_hg38_liftOver.bed",
        package = "xcore"
    )
)
GenomeInfoDb::seqlevels(roadmap_promoters, pruning.mode = "coarse") <-
    GenomeInfoDb::seqlevels(promoters_f5_core)
promoters_f5_core <- intersectGr(promoters_f5_core, roadmap_promoters)

# Select best promoter per gene
best_promoters <- GenomicRanges::mcols(promoters_f5_core) %>%
    as.data.frame() %>%
    dplyr::group_by(SYMBOL) %>%
    dplyr::slice(which.max(score)) %>%
    dplyr::pull(name)
promoters_f5_core <-
    promoters_f5_core[promoters_f5_core$name %in% best_promoters, ]

# restrict interaction matrices to core promoters only
remap_promoters_core <- remap_promoters[promoters_f5_core$name, ]
chip_atlas_promoters_core <- chip_atlas_promoters[promoters_f5_core$name, ]
```

This give us an interaction matrix with `r nrow(remap_promoters_core)` promoters
(rows) and `r ncol(remap_promoters_core)`, `r ncol(chip_atlas_promoters_core)` 
ChIP-seq experiments (columns) for ReMap2020, ChIP-Atlas respectively.

## Looking at promoter TF interaction matrix column wise (TF wise)

```{r tf_coverage, echo=FALSE}
# Functions
# Function calculates between columns correlation over sparse matrix
# Taken from https://stackoverflow.com/questions/5888287/running-cor-or-any-variant-over-a-sparse-matrix-in-r
sparse.cor <- function(x){
    n <- nrow(x)

    cMeans <- Matrix::colMeans(x)
    cSums <- Matrix::colSums(x)

    # Calculate the population covariance matrix.
    # There's no need to divide by (n-1) as the std. dev is also calculated the same way.
    # The code is optimized to minize use of memory and expensive operations
    covmat <- Matrix::tcrossprod(cMeans, (-2 * cSums + n * cMeans))
    crossp <- Matrix::as.matrix(Matrix::crossprod(x))
    covmat <- covmat + crossp

    sdvec <- sqrt(Matrix::diag(covmat)) # standard deviations of columns
    covmat / Matrix::crossprod(Matrix::t(sdvec)) # correlation matrix
}

# Data
srx2srastudy <- data.table::fread(
  file = system.file(
    "inst",
    "analysis",
    "srx2sraproject",
    "srx2study.csv",
    package = "xcore"
  )
)

translate <- function(x, key, value) {
  newx <- value[match(x, key)]
  newx
}

remap_tfs <- remap_promoters_core %>%
  colnames() %>%
  stringr::str_split(pattern = "\\.", n = 3) %>%
  do.call(what = rbind) %>%
  data.table::as.data.table()
colnames(remap_tfs) <- c("id", "tf", "background")
remap_tfs$coverage <- remap_promoters_core %>% Matrix::colSums()
remap_tfs$biotype <- remap_tfs$background %>% gsub(pattern = "_.*", replacement = "")
remap_tfs$condition <- remap_tfs$background %>% 
  sub(pattern = "(.*?)_", replacement = "")
remap_tfs$condition[! grepl("_", remap_tfs$background)] <- ""

# fix study ids
remap_tfs$study <- remap_tfs$id
remap_tfs$study[grepl(pattern = "ENCSR", x = remap_tfs$id)] <- 
  translate(grep(pattern = "ENCSR", x = remap_tfs$id, value = TRUE),
            srx2srastudy$Experiment,
            srx2srastudy$SRAStudy)
# restore ids
remap_tfs$id <- remap_promoters_core %>% colnames()

chip_atlas_tfs <- chip_atlas_promoters_core %>%
  colnames() %>%
  stringr::str_split(pattern = "_", n = 4) %>%
  do.call(what = rbind) %>%
  data.table::as.data.table()
colnames(chip_atlas_tfs) <- c("tf", "background", "biotype", "id")
chip_atlas_tfs$coverage <- chip_atlas_promoters_core %>% Matrix::colSums()

# fix study id
chip_atlas_tfs$study <- 
  srx2srastudy[match(chip_atlas_tfs$id, srx2srastudy$Experiment), ]$SRAStudy

# restore ids
chip_atlas_tfs$id <- chip_atlas_promoters_core %>% colnames()
```

There are `r sum(remap_tfs$coverage == 0)`, `r sum(chip_atlas_tfs$coverage == 0)`
zero coverage ChIP-seqs in ReMap2020, ChIP-Atlas respectively. I've removed 
TF with coverage below 11 from further analysis.

In ReMap2020 we focus only on the biotypes with >= 30 experiments excluding 
the "breast" and "liver" biotypes.

Moreover in ChIP-Atlas there is `r sum(chip_atlas_tfs$biotype == "Unclassified")`
with 'Unclassified' biotype, those are removed as well.

```{r chip_cleanup, echo=FALSE}
remap_tfs <- remap_tfs[remap_tfs$coverage > 10]

remap_tfs <- remap_tfs[condition == ""]

sel_biotypes <- remap_tfs$biotype %>% 
  table() %>% 
  `[`(. >= 30) %>% 
  names() %>%
  `[`(! . %in% c("breast", "liver"))
remap_tfs <- remap_tfs[biotype %in% sel_biotypes]


# annotate coverage
remap_tfs[, cov_type := cut(
  coverage,
  breaks = c(0, quantile(coverage, 0.25), quantile(coverage, 0.75), Inf),
  labels = c("low", "medium", "high")
)]

chip_atlas_tfs <- chip_atlas_tfs[chip_atlas_tfs$coverage != 0]
chip_atlas_tfs <- chip_atlas_tfs[chip_atlas_tfs$biotype != "Unclassified"]
```

```{r chip_seq_coverage, include=FALSE}
par(mfrow = c(1, 2), cex = 0.7, pty = "s")
coverageHist(x = remap_tfs$coverage,
             main = "ReMap2020 ChIP-seq coverage")
coverageHist(x = chip_atlas_tfs$coverage, 
             main = "ChIP-Atlas ChIP-seq coverage")

coverageHist(x = remap_tfs$tf %>% table() %>% as.numeric(),
             main = "ReMap2020 TFs")
coverageHist(x = chip_atlas_tfs$tf %>% table() %>% as.numeric(), 
             main = "ChIP-Atlas TFs")

coverageHist(x = remap_tfs$biotype %>% table() %>% as.numeric(),
             main = "ReMap2020 biotypes")
coverageHist(x = chip_atlas_tfs$biotype %>% table() %>% as.numeric(), 
             main = "ChIP-Atlas biotypes")
```

### TF / background confusion matrix

```{r tf_bg_confusion_matrix, echo=FALSE}
remap_tfs_tab <-
  data.table::dcast(data = remap_tfs,
                    formula = biotype ~ tf,
                    fun = length)
ord <- c(1, order(remap_tfs_tab[, -1] %>% colSums(), decreasing = TRUE) + 1)
data.table::setcolorder(remap_tfs_tab, ord)
remap_tfs_tab[, .r := -rowSums(remap_tfs_tab[, -1])]
data.table::setorder(remap_tfs_tab, .r)[, .r := NULL]

chip_atlas_tfs_tab <-
  data.table::dcast(data = chip_atlas_tfs,
                    formula = biotype ~ tf,
                    fun = length)
ord <- c(1, order(chip_atlas_tfs_tab[, -1] %>% colSums(), decreasing = TRUE) + 1)
data.table::setcolorder(chip_atlas_tfs_tab, ord)
chip_atlas_tfs_tab[, .r := -rowSums(chip_atlas_tfs_tab[, -1])]
data.table::setorder(chip_atlas_tfs_tab, .r)[, .r := NULL]

knitr::kable(remap_tfs_tab[, .(biotype, MYC, ELF1, MAX, YY1, RAD21, FOXA1, SIN3A)])
```
### Case study: select TP53, MYC other similar specific TFs and show heatmap
### with clustering, does they cluster by study or by experiment? -- towards
### collapsing method

```{r case_study_functions, echo=FALSE}
tfSummary <- function(sel_tf, meta) {
  print(meta[tf == sel_tf,
       .(
         n_id = length(id),
         n_study = length(unique(study)),
         n_biotype = length(unique(biotype))
       )])
  
  # print(meta[tf == sel_tf,
  #      hist(table(study), main = paste0(sel_tf, " study"))])
  # print(meta[tf == sel_tf,
  #      hist(table(biotype), main = paste0(sel_tf, " biotype"))])
}

getTFMat <- function(sel_tf, mat, meta) {
  i <- meta[tf == sel_tf, id] 
  mat[, i]
}

makeTFHeatmap <- function(tf_mat, meta, method = "pearson") {
  print(tf_mat %>% dim())
  
  # tf experiments correlation
  if (method == "binary") {
    tf_cor_d <- dist(Matrix::t(tf_mat), method = "binary") # binary
    tf_cor <- as.matrix(tf_cor_d)
  } else {
    tf_cor <- cor(tf_mat %>% as.matrix(), method = method)
    tf_cor_d <- as.dist(abs((1 - tf_cor) / 2))
  }
  
  # heatmap
  annotaion <- data.frame(
    study = meta$study,
    biotype = meta$biotype,
    coverage = meta$cov_type,
    row.names = meta$id,
    stringsAsFactors = TRUE
  )
  annotation_colors <- lapply(
    X = annotaion,
    FUN = function(x) {
      col <- stats::setNames(colorspace::qualitative_hcl(nlevels(x)), levels(x))
      # m <-
      #   unclass(x)[!duplicated(unclass(x)) &
      #                !duplicated(unclass(x), fromLast = TRUE)]
      # col[m] <- NA
      col
    }
  )
  labs <- paste0(meta$study, " ", meta$biotype)
  
  pheatmap::pheatmap(
    mat = tf_cor,
    scale = "none",
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    clustering_distance_rows = tf_cor_d,
    clustering_distance_cols = tf_cor_d,
    # fontsize = 5,
    # labels_col = labs,
    # labels_row = labs,
    # annotation_row = annotaion,
    annotation_col = annotaion,
    annotation_colors = annotation_colors,
    angle_col = 45,
    treeheight_col = 0
  )
}
```

### MYC

```{r myc_remap}
sel_tf <- "MYC"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

full_rows <- tf_mat_sub %>% Matrix::rowSums() %>% `==`(ncol(tf_mat_sub))
tf_mat_sub <- tf_mat_sub[! full_rows, ]

makeTFHeatmap(tf_mat = tf_mat_sub, 
              meta = remap_tfs[id %in% colsel])
```

### ELF1

```{r elf1_remap}
sel_tf <- "ELF1"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

makeTFHeatmap(tf_mat = tf_mat[rowsel, colsel], 
              meta = remap_tfs[id %in% colsel])
```

### MAX

```{r max_remap}
sel_tf <- "MAX"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

makeTFHeatmap(tf_mat = tf_mat[rowsel, colsel], 
              meta = remap_tfs[id %in% colsel])
```


### YY1

```{r yy1_remap}
sel_tf <- "YY1"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

makeTFHeatmap(tf_mat = tf_mat[rowsel, colsel], 
              meta = remap_tfs[id %in% colsel])
```

### RAD21

```{r rad21_remap}
sel_tf <- "RAD21"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

makeTFHeatmap(tf_mat = tf_mat[rowsel, colsel], 
              meta = remap_tfs[id %in% colsel])
```

### FOXA1

```{r foxa1_remap}
sel_tf <- "FOXA1"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

makeTFHeatmap(tf_mat = tf_mat[rowsel, colsel], 
              meta = remap_tfs[id %in% colsel])
```

### SIN3A

```{r sin3a_remap}
sel_tf <- "SIN3A"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

makeTFHeatmap(tf_mat = tf_mat[rowsel, colsel], 
              meta = remap_tfs[id %in% colsel])
```

### ESR1

```{r esr1_remap}
sel_tf <- "ESR1"
tf_mat <- getTFMat(sel_tf, remap_promoters_core, remap_tfs)

tfSummary(sel_tf, remap_tfs)

# coverageHist(Matrix::colSums(tf_mat), main = paste0(sel_tf, " coverage"))
# coverageHist(Matrix::rowSums(tf_mat), main = paste0(sel_tf, " rowSums"))


# remove experiments with coverage below 10 and promoters with coverage lower than 3
# look only at biotypes with at least 2 replicates
colsum_id <- colnames(tf_mat)[Matrix::colSums(tf_mat) > 10]
colsel <- remap_tfs[id %in% colsum_id, .(id, .N), by = "biotype"][N >= 2, id]
rowsel <- Matrix::rowSums(tf_mat[, colsel]) >= 2

tf_mat_sub <- tf_mat[rowsel, colsel]

makeTFHeatmap(tf_mat = tf_mat[rowsel, colsel], 
              meta = remap_tfs[id %in% colsel])
```

### Correlation how it works?

```{r explore_correlation}
sel_tf <- remap_tfs_tab[, -1][, sapply(.SD, sum) %>% `[`(. >= 10)] %>% names()
m_biotypes <- remap_tfs_tab[, sel_tf, with = FALSE] %>% rowSums() %>% `>=`(10)
sel_biotypes <- remap_tfs_tab[m_biotypes, biotype]

sel_id <- remap_tfs[biotype %in% sel_biotypes & tf %in% sel_tf, id]
tf_mat <- remap_promoters_core[, sel_id]

tf_mat_cor <- tf_mat %>% as.matrix() %>% cor()

# same tf
same_tf_cor <- sapply(
  X = sel_tf,
  FUN = function(x) {
    m_row <- grep(x, rownames(tf_mat_cor), value = TRUE)
    m_col <- grep(x, colnames(tf_mat_cor), value = TRUE)
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

# diff tf
diff_tf_cor <- sapply(
  X = sel_tf,
  FUN = function(x) {
    m_row <- grep(x, rownames(tf_mat_cor), value = TRUE)
    m_col <- grep(x, colnames(tf_mat_cor), value = TRUE, invert = TRUE)
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

boxplot(unlist(same_tf_cor), unlist(diff_tf_cor), 
        names = c("same TF", "different TFs"), 
        main = "Correlation between experiments")

p1 <- hist(unlist(same_tf_cor), freq = FALSE, breaks = 50)                     # centered at 4
p2 <- hist(unlist(diff_tf_cor), freq = FALSE, breaks = 50)                     # centered at 6
plot(p1, col=rgb(0,0,1,1/4), xlim=c(-0.4, 1), ylim=c(0, 4), freq = FALSE)  # first histogram
plot(p2, col=rgb(1,0,0,1/4), xlim=c(-0.4, 1), freq = FALSE, add=T)

th <- seq(0, 1, by = 0.01)
crit <- sapply(th, function(th) sum(unlist(same_tf_cor) > th) / sum(unlist(diff_tf_cor) > th))

# same biotype
same_bio_cor <- sapply(
  X = sel_biotypes,
  FUN = function(x) {
    m_row <- grep(x, rownames(tf_mat_cor), value = TRUE)
    m_col <- grep(x, colnames(tf_mat_cor), value = TRUE)
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

# diff biotype
diff_bio_cor <- sapply(
  X = sel_biotypes,
  FUN = function(x) {
    m_row <- grep(x, rownames(tf_mat_cor), value = TRUE)
    m_col <- grep(x, colnames(tf_mat_cor), value = TRUE, invert = TRUE)
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

boxplot(unlist(same_bio_cor), unlist(diff_bio_cor), 
        names = c("same biotype", "different biotype"), 
        main = "Correlation between experiments")

# same study
same_study_cor <- sapply(
  X = remap_tfs[, unique(study)],
  FUN = function(x) {
    m_row <- grep(x, rownames(tf_mat_cor), value = TRUE)
    m_col <- grep(x, colnames(tf_mat_cor), value = TRUE)
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

# diff study
diff_sutdy_cor <- sapply(
  X = remap_tfs[, unique(study)],
  FUN = function(x) {
    m_row <- grep(x, rownames(tf_mat_cor), value = TRUE)
    m_col <- grep(x, colnames(tf_mat_cor), value = TRUE, invert = TRUE)
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

boxplot(unlist(same_study_cor), unlist(diff_sutdy_cor), 
        names = c("same study", "different study"), 
        main = "Correlation between experiments")

remap_tfs[, .N, by = .(study, tf)]$N %>% summary()


# TF classes
df <- data.table::fread(system.file("inst", "analysis", "cis_bp_tf_class.txt", package = "xcore"))
df <-
  df[, .(TF_Name,
         Family_Name = unique(Family_Name),
         DBDs = unique(DBDs)), 
     by = TF_Name]
df <- remap_tfs[df, on = c("tf" = "TF_Name")]
df <- df[! is.na(id)]

# same tf class
same_class_cor <- sapply(
  X = unique(df$Family_Name),
  FUN = function(x) {
    m_row <- rownames(tf_mat_cor) %in% df[Family_Name == x, id]
    m_col <- colnames(tf_mat_cor) %in% df[Family_Name == x, id]
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

# diff tf class
diff_class_cor <- sapply(
  X = unique(df$Family_Name),
  FUN = function(x) {
    m_row <- rownames(tf_mat_cor) %in% df[Family_Name == x, id]
    m_col <- colnames(tf_mat_cor) %in% df[Family_Name != x, id]
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

boxplot(unlist(same_class_cor), unlist(diff_class_cor), 
        names = c("same family", "different family"), 
        main = "Correlation between experiments")

# coverage dependence
# same tf high cov
same_tf_cor_high <- sapply(
  X = sel_tf,
  FUN = function(x) {
    m_row <- rownames(tf_mat_cor) %in% remap_tfs[tf == x & cov_type == "high", id]
    m_col <- colnames(tf_mat_cor) %in% remap_tfs[tf == x & cov_type == "high", id]
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

# same tf medium cov
same_tf_cor_medium <- sapply(
  X = sel_tf,
  FUN = function(x) {
    m_row <- rownames(tf_mat_cor) %in% remap_tfs[tf == x & cov_type == "medium", id]
    m_col <- colnames(tf_mat_cor) %in% remap_tfs[tf == x & cov_type == "medium", id]
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

# same tf low cov
same_tf_cor_low <- sapply(
  X = sel_tf,
  FUN = function(x) {
    m_row <- rownames(tf_mat_cor) %in% remap_tfs[tf == x & cov_type == "low", id]
    m_col <- colnames(tf_mat_cor) %in% remap_tfs[tf == x & cov_type == "low", id]
    sel_cor_mat <- tf_mat_cor[m_row, m_col]
    sel_cor_mat[upper.tri(sel_cor_mat, diag = FALSE)]
  }
)

boxplot(
  unlist(same_tf_cor_high),
  unlist(same_tf_cor_medium),
  unlist(same_tf_cor_low),
  names = c("high coverage", "medium coverage", "low coverage"),
  main = "Correlation between experiments, same TF"
)
```

# Multiple Correspondence Analysis

```{r}
X <- tf_mat %>% 
  as.matrix() %>%
  t() %>%
  as.data.frame() %>% 
  rapply(f = as.character, classes = "numeric", how = "replace")
# X <- cbind(X, remap_tfs[match(rownames(X), id), .(id, biotype, tf, coverage)])

res.mca <- FactoMineR::MCA(
  X = X,
  ncp = 5)

factoextra::fviz_screeplot(res.mca, addlabels = TRUE, ylim = c(0, 45))

factoextra::fviz_mca_ind(res.mca, 
             label = "none", # hide individual labels
             habillage = remap_tfs[match(rownames(X), id), factor(tf)], # color by groups 
             #palette = c("#00AFBB", "#E7B800"),
             # addEllipses = TRUE, ellipse.type = "confidence",
             ggtheme = ggplot2::theme_minimal()) 
```

# Association rules

```{r}
core_transactions_remap <- arules::transactions(t(remap_promoters_core))
```

# TF tissue specificity

Here I will define TF tissue specificity as an average tau of the promoters it
intersect with.

```{r}
tf_tau <- vapply(X = colnames(remap_promoters_core), 
                 FUN = function(x) {
                   i <- remap_promoters_core[, x, drop = T] != 0
                   ids <- rownames(remap_promoters_core)[i]
                   promoters_f5_expression[`00Annotation` %in% ids, mean(tau)]
                 },
                 FUN.VALUE = numeric(1L))
remap_tfs[, tau := tf_tau[id]]
```